---
title: "Rome_Taxi_data"
author: "Lam"
date: "2023-12-05"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overall 

The analysis of results from urban mobility simulations can provide very valuable information for
the identification and addressing of problems in an urban road network. Public transport vehicles such as buses
and taxis are often equipped with GPS location devices and the location data is submitted to a central server
for analysis. The metropolitan city of Rome, Italy collected location data from 320 taxi drivers that work in the centre of Rome. Data was collected during the period from 01/Feb/2014 until 02/March/2014. An extract of the
dataset is found in taxi.csv. The dataset contains 4 attributes:
      1. ID of a taxi driver. This is a unique numeric ID.
      2. Date and time in the format Y:m:d H:m:s.msec+tz, where msec is microseconds, and tz is a
      time-zone adjustment. 
      3. Latitude
      4. Longitude
      
For a further description of this dataset: https://ieee-dataport.org/open-access/crawdad-romataxi


```{r Rome Taxi}
taxi <- read.csv("taxi.csv")
summary(taxi)
```

## Purpose of the project 
Perform a general analysis of this dataset. Learn to work with large datasets. Obtain general information about the
behaviour of some taxi drivers. Analyse and interpret results.

## Specific requirement 
(a) Plot the location points (2D plot using all of the latitude, and longitude value pairs in the dataset). 
Indicate invalid points, outliers or noise points. 

(b) Compute the minimum, maximum, and mean location values.

(c) Obtain the most active, least active, and average activity of the taxi drivers (most time driven, least
time driven, and meantime driven). Explain the rationale of my approach and results.

## Pre-processing 
Load all the packages needed 
```{r Load Packages}
  library (ggplot2)
  library (tidyverse)
  library (maps)
  library (lubridate)
```
Plot the location points of all taxi drivers 
```{r Plot all taxi Driver} 
subset_world_map <- subset(map_data("world"), lat >35 & lat <55 & long >-1 & long <20)
ggplot(taxi, aes(x = Longitude, y = Latitude)) +
  ggtitle("Location of Taix Drivers in Rome from 1st Feb 2014 to 2nd Mar 2014") +
  geom_polygon(data = subset_world_map, aes(x = long, y = lat, group = group), fill = "lightgrey", color = "black")+
  geom_point(size = 1, color = "blue")+
  xlab("Longitude")+
  ylab("Latitude")
```

We can see that there 2 regions of outliners, that were outside Rome. So I decided to move these outliners
``` {r remove outliners}

  taxi_clean <- subset(taxi, taxi$Latitude <45 & taxi$Longitude <15)
  taxi_clean$Date.and.Time <- ymd_hms(taxi_clean$Date.and.Time)
summary (taxi_clean)
```
There is another problem, the Date&Time is in Char type, so I need to convert it to time format, using 'ymd_hms()' from package 'lubridate'

## Analysis 
First we need to group 'taxi_clean' by 'DriveNo'
```{r group by DriveNo}
taxi_grouped <- taxi_clean %>% group_by(DriveNo)
summary(taxi_grouped)

```

Now I calculate the total running time 
``` {r Total_time}
taxi_total_time <- taxi_grouped %>% 
                    arrange(DriveNo, Date.and.Time) %>% 
                    mutate(time_Diff = difftime(lead(Date.and.Time), Date.and.Time)) %>%
                    summarise(total_Time = sum(time_Diff, na.rm = TRUE))

taxi_total_time$total_Time <- as.numeric(taxi_total_time$total_Time)

avg_time_driven <- mean(taxi_total_time$total_Time)/3600/24
most_time_driven <- max(taxi_total_time$total_Time)/3600/24
least_time_driven <- min(taxi_total_time$total_Time)/3600/24
print(avg_time_driven)
print(most_time_driven)
print(least_time_driven)

```
According, the average time driven from 1st Feb to 2nd Mar is approximately 26.35 days and the least time is 0.05days or appoximately 1.12 hours and the most time driven is 29 days
